---
title: "Whole-Slide Pathology Images: Binary Pipelines Across Environments"
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
library(condathis)
```

## The Problem

Whole-slide images (WSI) used in digital pathology, files like Aperio `.svs`,
Hamamatsu `.ndpi`, or MIRAX `.mrxs`, are not ordinary images.
A single slide is a multi-resolution pyramid of tiled, compressed pixels that
can be tens of gigapixels in size.
Reading them requires the OpenSlide C library, and in practice the only
convenient, scriptable way to pull a region out of a slide is through a tool
built on top of it, most commonly a Python environment
(`openslide-python`) or `libvips` compiled with OpenSlide support.

That creates a real dependency problem for an R user:

* Installing a Python/OpenSlide stack just to read one file format pollutes
  (or conflicts with) whatever environment your actual analysis lives in.
* The tool you'd use to *look at* the extracted region, an image
  processing library like ImageMagick, often needs its own, different set of
  compiled dependencies (its own `libjpeg`, `libpng`, `libtiff`...), which can
  clash with the versions OpenSlide/`libvips` need.
* None of this is reproducible unless the exact environments are pinned
  somewhere.

This article uses `condathis` to solve all three: two CLI tools, each
completely isolated in its own Conda environment, connected by a real Unix
pipe, streaming binary image data from one environment straight into the
other, and finally back into R for plotting. Nothing is installed system-wide
and nothing leaks between environments.

## The Slide

We'll use a real, public test slide from the OpenSlide project: a small
region cropped from an Aperio `.svs` file, about 2 MB.

```{r}
#| label: download-slide
data_dir <- fs::path("data-raw")
fs::dir_create(data_dir)

svs_path <- fs::path(data_dir, "CMU-1-Small-Region.svs")

if (isFALSE(fs::file_exists(svs_path))) {
  utils::download.file(
    url = paste0(
      "https://openslide.cs.cmu.edu/download/",
      "openslide-testdata/Aperio/CMU-1-Small-Region.svs"
    ),
    destfile = svs_path,
    mode = "wb",
    quiet = TRUE
  )
}

fs::file_size(svs_path)
```

## Two Isolated Toolchains

We need two tools that have no business sharing a Conda environment:

1. **`libvips`**, built with OpenSlide support, to open the `.svs` pyramid
   and extract pixels from it.
2. **ImageMagick**, to post-process the extracted region.

```{r}
#| label: create-envs
create_env(
  packages = "conda-forge::libvips",
  channels = "conda-forge",
  env_name = "wsi-vips-env",
  verbose = "silent"
)

create_env(
  packages = "conda-forge::imagemagick",
  channels = "conda-forge",
  env_name = "wsi-magick-env",
  verbose = "silent"
)
```

Each `create_env()` call resolves and installs its own dependency graph.
`libvips` pulls in OpenSlide, `openjpeg`, and friends; ImageMagick pulls in
its own delegate libraries. They never see each other.

## Orienting Ourselves: a Thumbnail

Before extracting a region, let's generate a small overview of the whole
slide with a single `run()` call, `vips thumbnail` scales the pyramid down
directly, it never decodes full-resolution pixels.

```{r}
#| label: thumbnail
thumb_png <- fs::file_temp(ext = "png")

run(
  "vips", "thumbnail", svs_path, thumb_png, "400",
  env_name = "wsi-vips-env",
  verbose = "silent"
)

thumb_img <- png::readPNG(thumb_png)
dim(thumb_img)
```

## Building the Pipeline

Now the actual pipeline: `run_pipeline()` chains two commands, in two
different Conda environments, connected by a kernel-level pipe managed
entirely by the child processes. The PNG bytes produced by `vips crop` never
pass through R, they go straight into ImageMagick's stdin.

```{r}
#| label: run-pipeline
region_png <- fs::file_temp(ext = "png")

# left, top, width, height (pixels) of the region to crop out of the slide
crop_box <- c(800, 1600, 800, 800)

pipeline_res <- run_pipeline(
  cmds = list(
    list(
      cmd = c("vips", "crop", svs_path, ".png", as.character(crop_box)),
      env_name = "wsi-vips-env"
    ),
    list(
      cmd = c(
        "magick", "-",
        "-modulate", "100,150,100", # boost saturation
        "-sharpen", "0x1",
        "png:-"
      ),
      env_name = "wsi-magick-env",
      stdout = region_png
    )
  )
)

pipeline_res
```

Two details are load-bearing here:

* **`stdin`/`-`/`png:-`**: the middle of the pipe is fully binary and never
  touches R, `processx` connects the two child processes' file descriptors
  directly. Only the *first* command's stdin and the *last* command's stdout
  are ever visible to R, everything in between is a raw byte stream.
* **`stdout = region_png`** on the last command: `run_pipeline()`'s default
  `stdout = "|"` capture decodes output as UTF-8 text, which would corrupt
  binary PNG bytes. Since the final artifact here is a binary image, we
  redirect it straight to a file instead of capturing it as a character
  string.

## Plotting the Result

With the pipeline's output on disk, reading it into R is a normal
`png::readPNG()` call.

```{r}
#| label: plot-result
#| fig-width: 9
#| fig-height: 5
region_img <- png::readPNG(region_png)

th <- dim(thumb_img)[1]
tw <- dim(thumb_img)[2]
scale_x <- tw / 2220
scale_y <- th / 2967
rx0 <- crop_box[1] * scale_x
ry0 <- crop_box[2] * scale_y
rx1 <- (crop_box[1] + crop_box[3]) * scale_x
ry1 <- (crop_box[2] + crop_box[4]) * scale_y

withr::local_par(mfrow = c(1, 2), mar = c(1, 1, 2, 1))

plot(
  c(0, tw), c(0, th),
  type = "n", xlab = "", ylab = "", asp = 1, axes = FALSE,
  main = "Whole slide (thumbnail)"
)
rasterImage(thumb_img, 0, 0, tw, th)
rect(rx0, th - ry1, rx1, th - ry0, border = "red", lwd = 2)

plot(
  c(0, crop_box[3]), c(0, crop_box[4]),
  type = "n", xlab = "", ylab = "", asp = 1, axes = FALSE,
  main = "Extracted + enhanced region"
)
rasterImage(region_img, 0, 0, crop_box[3], crop_box[4])
```

The red box on the thumbnail marks exactly the region `libvips` cropped out
of the pyramid, ImageMagick then boosted its saturation and sharpened it, all
without R ever holding a raw pixel buffer for anything but the final,
already-small PNG.

## Binary Data In: Raw Bytes as stdin

The pipeline above only needed a *file* as its output. `run()` and
`run_pipeline()` also accept raw binary data as *input*, written directly to
a process's stdin. To show it, let's read the region PNG we just produced
back into R as a raw vector and pipe those bytes into ImageMagick's `identify`
without writing anything new to disk:

```{r}
#| label: raw-input
png_bytes <- readBin(region_png, what = "raw", n = fs::file_size(region_png))
length(png_bytes)

identify_res <- run(
  "magick", "identify", "-",
  env_name = "wsi-magick-env",
  stdin = "|",
  input = png_bytes,
  verbose = "silent"
)

parse_output(identify_res)
```

`input` accepts a character or raw vector, here we pass the raw PNG bytes
already sitting in R's memory straight into the child process, no temporary
file involved.

## Cleaning Up

```{r}
#| label: cleanup
remove_env("wsi-vips-env")
remove_env("wsi-magick-env")
```

## Why This Matters

* **Isolation**: `libvips`+OpenSlide and ImageMagick never share a
  dependency graph, so there's no solver conflict to fight and each can be
  upgraded independently.
* **True binary pipes**: data moves between the two Conda environments as
  raw bytes over a kernel pipe, exactly like a shell `|`, R is only on the
  hook for the first command's stdin and the last command's stdout.
* **Reproducibility**: the whole toolchain, from slide to plotted region, is
  defined in this article. Anyone who runs it gets the same environments and
  the same pixels.
